The Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
some extensions, do the XXXs

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

\title{The Way Forward in Parallelizing Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

@@

%% \paragraph{dynamic language VM problems}
%% XXX:
%% - high allocation rate (short lived objects)\\
%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in the interpreter. The basic guarantee is that the GIL
may only be released in-between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter.
[xxx mention that the interpreter is typically
  very large and maintained by open-source communities]

@@

While one can argue that STM requires the insertion of read and write
barriers in the whole program, this can be done automatically and
locally by a program transformation\cite{felber07}. There are attempts
to do the same for fine-grained locking\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable.
The effectiveness of these approaches still has to be proven for our
use case.
[xxx or maybe: "The effectiveness of these approaches is doubtful in our
use case --- for example, it makes it close to impossible to order the
locks consistently or to know in advance which locks a transaction will
need.]

>>>>

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

\title{The Way Forward in Parallelising Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\begin{abstract}
  Dynamic languages became very popular in recent years. At some
  point, the need for concurrency arised, and many of them made the
  choice to use a single global interpreter lock (GIL) to synchronize
  the interpreter in a multithreading scenario. This choice however
  makes it impossible to actually run code in parallel.

  Here we want to compare different approaches to replacing the GIL
  with a technology that allows parallel execution. We look at
  fine-grained locking, shared-nothing, and transactional memory (TM)
  approaches. We argue that software-based TM systems are the most
  promising, especially since they also enable the introduction of
  atomic blocks as a better synchronization mechanism in the language.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

@@

In this section we examine the approaches and highlight their
advantages and disadvantages.
%% \paragraph{dynamic language VM problems}
%% XXX:
%% - high allocation rate (short lived objects)\\
%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in the interpreter. The basic guarantee is that the GIL
may only be released in-between bytecode instructions. The interpreter
can thus rely on complete isolation and atomicity of these
instructions. Additionally, it provides the application with a
sequential consistency model\cite{lamport79}. As a consequence,
applications can rely on certain operations to be atomic and that they
will always be executed in the order in which they appear in the
code. While depending on this may not always be a good idea, it is
done in practice. A GIL-replacement should therefore uphold these
guarantees, while preferably also be as easily implementable as a GIL
for the interpreter. The latter can be especially important since
many of these languages are developed and maintained by very large
open-source communities, which are not easy to coordinate.

@@

While one can argue that STM requires the insertion of read and write
barriers in the whole program, this can be done automatically and
locally by a program transformation\cite{felber07}. There are attempts
to do the same for fine-grained locking\cite{bill06} but they require
a whole program analysis since locks are inherently non-composable.
The effectiveness of these approaches is doubtful in our use case,
since we execute bytecode instructions in any order defined by a
script only known at runtime. This makes it close to impossible to
order locks consistently or to know in advance which locks a
transaction will need.
